---
title: "`ctsframework` : A Modular Framework for Performing Clinical Trial Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`ctsframework` : A Modular Framework for Performing Clinical Trial Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval=TRUE, message=FALSE, warning=FALSE, error=FALSE)

# Create a temporary directory to hold the results
cts_results_path <- tempfile()
dir.create(cts_results_path, FALSE, TRUE)
knitr::opts_knit$set(root.dir=cts_results_path)
```

```{r, echo=FALSE}
#devtools::load_all()
library(ctsframework)

# Reset all counters
reset_counter(.all=TRUE)
```


## Introduction

One of the core elements of pharmacometrics is to perform clinical trial
simulations (CTS) to help inform decisions in clinical drug development (for
instance, to explore different dosing regimens). In most cases, R is a suitable
environment for this, and there are many excellent packages that can be
leveraged.  In practice, what tends to happen is that there is a desire to
explore a large number of "what if" scenarios, either different treatments,
populations, models or other tweaked parameters, and as a consequence the
resulting code tends to grow in complexity and become cumbersome to manage. The
purpose of the `ctsframework` package is to make the CTS process more modular,
and to help manage the complexity.

Much as the infamous `ggplot2` package is designed around a "grammar of
graphics", this package offers something like a "grammar" of semantically named
functions that correspond to different components of a clinical trial
simulation.  Designed for maximum flexibility and convenience. The central
object is the `scenario`, which is composed of specific "modules" corresponding
to different aspects of CTS.

Withing `ctsframework`, a CTS is specified by one or more scenarios. A scenario
is broken down into multiple steps or "modules":

- **parameters**: This module can be used to specify any general parameters
  that are related to a scenario. An example could be the sample size, for
  instance.

- **population**: In most clinical trial simulations, the starting point is
  defining a population of 'virtual patients'. This module is where you would
  specify the distribution of demographic characteristics, implement
  inclusion/exclusion criteria and so on.

- **subject_model**: I don't usually like referring to people as 'subjects',
  but this is the common terminology used in clinical trials. This is the place
  where you could specify subject-level differences, like covariate effects or
  between-subject variability, and individual-level parameters.

- **treatment**: In most clinical trial simulations, the treatment will be of
  primary interest. This module is intended for specifying treatments, which
  can be a dose, formulation etc.

- **pkpd_model**: In pharmacometric applications, there will typically be a
  PK/PD model consisting of differential equations, which can be specified
  in this module.

- **simulation**: At this point, it's time to carry out the actual simulation.
  This module contains the code that runs it.

- **endpoints**: From the simulation output (which could for instance be a PK
  concentration-time profile, or some other complex object), a number of
  endpoints can be derived, which is what this module is intended for.

- **summarization**: This module is intended for summarizing results across
  individuals, such as taking means, medians, specific quantiles, etc. (Note:
  this module can only be used to summarize results within a scenario, not
  across scenarios; the latter is the purpose of the `collate()` function,
  described below).

In fact, each module is very general, and essentially equivalent; each can
contain any number of named items of arbitrary complexity. The semantic module
names are simply there to help make the code more organized and intentional.
All the modules are optional.

The power of the framework is that each module subitem can be overridden to
creat a new scenario. At each step, the submitems of the modules that appear
higher up are available and can be used in the subsequent computation. By
mixing and matching different modules, it is easy to specify many different
scenarios without a lot of code repetition.

## A complete pharmacometrics example

For this example we will use the following packages, so load them first:

```{r}
library(data.table)
library(mrgsolve)
library(linpk)
library(mappings)
library(ggplot2)
library(table1)
library(ttt)
```

Also, for reproducibility we had better set the RNG seed:

```{r}
set.seed(2755139)
```


### Defining a base scenario

Now, we begin by defining our first scenario, which will be our "base" scenario
(latter we will see how we can derive new scenarios from this one). The example
will be to simulate from a simple 2-compartment population PK model (it could
be a simplified monoclonal antibody model). This code block will apepar quite
long, but we will go through and explain each elements
afterwards.

```{r}
base_scenario <- scenario(name="Base Scenario",

    parameters = parameters(
        n  = 1000,
        th = list(CL=0.0194, V1=4.51, Q=0.0435, V2=2.82, CL_WT=0.46, V1_WT=0.87),
        om = LTmat(c(0.1, 0.01, 0.1), .names=c("nCL", "nV1"))
    ),

    population = population(
        wtcatfn = cut_mapping(c(0, 60, 100, Inf), c("<60 kg", "60 to <100 kg", ">= 100 kg"), right=TRUE),
        {
            data.table(ID=1:n, wt=exp(rnorm(n, log(74), 0.2)))[, wtcat := wtcatfn(wt)]
        }
    ),

    subject_model = subject_model(
        {
            eta <- generateETA(n, omega=om)

            cbind(population, eta)[, `:=`(
                CL = th$CL * exp(nCL) * (wt/70)^th$CL_WT,
                V1 = th$V1 * exp(nV1) * (wt/70)^th$V1_WT
            )]
        }
    ),

    treatment = treatment(
        {
            population[, .(ID, amt=wt*3, tinf=0.5, addl=5, ii=2*24*7)]
        }
    ),

    pkpd_model = pkpd_model(
        {
            model.code <- '
            [param]   CL=1, V1=1, V2=1, Q=1
            [pkmodel] cmt="CENTRAL PERIPH"
            [table]   capture conc = CENTRAL/V1;
            '

            mcode_cache("2cmt", model.code)
        }
    ),

    simulation = simulation(
        {
            eoi <- realize_addl(as.ev(treatment)) |> as.data.frame() |> with(time + (amt/rate)) |> unique() |> sort()

            sim <- pkpd_model |>
                param(th) |>
                idata_set(subject_model) |>
                ev(as.ev(treatment)) |>
                Req(conc) |>
                update(end=12*24*7, delta=24, add=eoi) |>
                mrgsim()

            simdat <- sim |> as.data.frame() |> as.data.table()

            population[simdat, on="ID"]
        }
    ),

    endpoints = endpoints(
        {
            auc.by.trapezoid <- function(x, y) sum(0.5*diff(x)*(y[-1] + y[-length(y)]))

            simulation[time >= 6*24*7, .(
                cmin=min(conc),
                cmax=max(conc),
                cavg=auc.by.trapezoid(time, conc)/diff(range(time))),
            by=ID][population, on="ID"]
        },

        outputs = csv_output(endpoints, name="sim-exposure")
    ),

    summarization = summarization(
        {
            list(
                overall = simulation[, qi(conc, level=0.9), by=time],
                bywtcat = simulation[, qi(conc, level=0.9), by=.(time, wtcat=wtcatfn(wt))]
            )
        },

        outputs = output_list(
            graphical_output(name="sim-profile-90pctPI-overall", width=7, height=5, {
                ggplot(summarization$overall, aes(x=time/(24*7), y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (weeks)", y="Concentration (\U{00B5}g/mL)") +
                    scale_x_continuous(breaks=seq(0, 12, 2)) +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            }),
            graphical_output(name="sim-profile-90pctPI-bywtcat", width=7, height=5, {
                ggplot(summarization$bywtcat, aes(x=time/(24*7), y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (weeks)", y="Concentration (\U{00B5}g/mL)") +
                    scale_x_continuous(breaks=seq(0, 12, 2)) +
                    facet_wrap(~ wtcat) +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            })
        )
    )
)
```

Now, we will go through the above code and briefly explain each module separately.

We start by looking a the `parameters`

```{r, eval=F}
    parameters = parameters(
        n  = 1000,
        th = list(CL=0.0194, V1=4.51, Q=0.0435, V2=2.82, CL_WT=0.46, V1_WT=0.87),
        om = LTmat(c(0.1, 0.01, 0.1), .names=c("nCL", "nV1"))
    )
```

We have specified a 3 parameters: the number of individual we want to simulate
(n), a THETA vector of fixed-effect parameters (th), and an OMEGA matrix for
between-individual variability (om). These parameters will be used in
subsequent code. Since each module is completely general, we are free to
specify whatever we want, and whatever help to make the code more organized.

```{r, eval=F}
    population = population(
        wtcatfn = cut_mapping(c(0, 60, 100, Inf), c("<60 kg", "60 to <100 kg", ">= 100 kg"), right=TRUE),
        {
            data.table(ID=1:n, wt=exp(rnorm(n, log(74), 0.2)))[, wtcat := wtcatfn(wt)]
        }
    )
```

There are actually 2 things being specified here. One is a named function
`wtcatfn()` that transforms body weight into weight categories. The other is an
unnamed code block that generates a population dataset. Note that the parameter
`n` appears in this code block, as does the `wtcatfn()` defined previously.
Again, anything that appear higher up is accessible in the code. However, it is
important to note that an unnamed code block such as we have used here can be
referred in subsequent modules using the module name in which is was specified
(in this case, `population`).

```{r, eval=F}
    subject_model = subject_model(
        {
            eta <- generateETA(n, omega=om)

            cbind(population, eta)[, `:=`(
                CL = th$CL * exp(nCL) * (wt/70)^th$CL_WT,
                V1 = th$V1 * exp(nV1) * (wt/70)^th$V1_WT
            )]
        }
    )
```

In an unnamed code block, we use the augment the population dataset with
individual level random effects and parameters.

```{r, eval=F}
    treatment = treatment(
        {
            population[, .(ID, amt=wt*3, tinf=0.5, addl=5, ii=2*24*7)]
        }
    )
```

For the treatment, we will assume a dose of 3 mg/kg given as a half-hour
infusion every two weeks over 12 weeks.

```{r, eval=F}
    pkpd_model = pkpd_model(
        {
            model.code <- '
            [param]   CL=1, V1=1, V2=1, Q=1
            [pkmodel] cmt="CENTRAL PERIPH"
            [table]   capture conc = CENTRAL/V1;
            '

            mcode_cache("2cmt", model.code)
        }
    )
```

In an unnamed code block, the 2-compartment PK model is defined using
`mrgsolve` syntax (the `mrgsolve` package has great documentation). The result
is the compiled model.

```{r, eval=F}
    simulation = simulation(
        {
            eoi <- realize_addl(as.ev(treatment)) |> as.data.frame() |> with(time + (amt/rate)) |> unique() |> sort()

            sim <- pkpd_model |>
                param(th) |>
                idata_set(subject_model) |>
                ev(as.ev(treatment)) |>
                Req(conc) |>
                update(end=12*24*7, delta=24, add=eoi) |>
                mrgsim()

            simdat <- sim |> as.data.frame() |> as.data.table()

            population[simdat, on="ID"]
        }
    )
```

In an unnamed code block, the concentration-time profiles are generated using
`mrgsolve`.

```{r, eval=F}
    endpoints = endpoints(
        {
            auc.by.trapezoid <- function(x, y) sum(0.5*diff(x)*(y[-1] + y[-length(y)]))

            simulation[time >= 6*24*7, .(
                cmin=min(conc),
                cmax=max(conc),
                cavg=auc.by.trapezoid(time, conc)/diff(range(time))),
            by=ID][population, on="ID"]
        },

        outputs = csv_output(endpoints, name="sim-exposure")
    )
```

Next, we derive some "endpoints", in this case exposure metrics over the last
treament interval, for each individual. For the first time, a module contains
an output -- the resulting dataframe will be stored in a CSV file. The location
of this CSV file will be explained later.

```{r, eval=F}
    summarization = summarization(
        {
            list(
                overall = simulation[, qi(conc, level=0.9), by=time],
                bywtcat = simulation[, qi(conc, level=0.9), by=.(time, wtcat=wtcatfn(wt))]
            )
        },

        outputs = output_list(
            graphical_output(name="sim-profile-90pctPI-overall", width=7, height=5, {
                ggplot(summarization$overall, aes(x=time, y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (h)", y="Concentration (\U{00B5}g/mL)") +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            }),
            graphical_output(name="sim-profile-90pctPI-bywtcat", width=7, height=5, {
                ggplot(summarization$bywtcat, aes(x=time, y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (h)", y="Concentration (\U{00B5}g/mL)") +
                    facet_wrap(~ wtcat) +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            })
        )
    )
```

We summarize the PK concentration-time profiles by taking the median, 5th and
95th percentiles at each time point. We do this both overall, and stratified by
weight category.  Then, we produce two graphical outputs using `ggplot`.


### Running a scenario

To run the base scenario, we do the following:

```{r, eval=F}
CTS(base_scenario)
```

After the scenario has finished running, there is a new folder in the current
directory named "`r file.path(".", "cts_results", as.Date(Sys.time()), "Base
Scenario")`" which contains the specified outputs each in their own subfolder
corresponding to the name of the module where it was specified. For instance,
the subfolder named `summarization` contains these 2 plots:

```{r, eval=T, echo=F, fig.width=7, fig.height=5}
x <- CTS(base_scenario, .path=cts_results_path)
outputs <- attr(x$scenario$summarization, "outputs", exact=TRUE)
outres <- rlang::eval_tidy(outputs, x)
outres[[1]]
outres[[2]]
```

The way the folder structure is defined is as follows. When the `CTS()`
function is called, a `.path` argument specifies the root path for the
simulation; the defaut is `cts_results/dd-mm-yyyy`, where `dd-mm-yyyy` is the
current date. Under this root folder, there will be a subfolder for each
scenario, named according to the scenario name (each scenario has a name
attribute, the default being `Scenario` followed by a sequential number). Then,
under each scenario folder there may be subfolders corresponding to individual
modules. This folder structure helps to keep things organized neatly.

The simulation root folder also contains a file called `scenarios.csv` that
contains useful information about each scenario that has been run.

### Deriving a new scenario from an existing one

Often there is more that one scenario to simulate, and here's where it starts
to get interesting. We can derive a new scenario from an existing one by
essentially specifying only the parts that we want to override.

Here, we will derive two new scenarios from the base scenario to explore dosing
the is independent of body weight (recall that the base regimen was 3 mg/kg
Q2W): 200 mg Q2W, and 300 mg Q3W.

```{r, eval=F}
scenario_fixed_200mg_q2w <- derive_scenario(base_scenario,
    treatment = treatment(name="200 mg Q2W", {
        population[, .(ID, amt=200, tinf=0.5, addl=5, ii=2*24*7)]
    })
)

scenario_fixed_300mg_q3w <- derive_scenario(base_scenario,
    treatment = treatment(name="300 mg Q3W", {
        population[, .(ID, amt=300, tinf=0.5, addl=3, ii=3*24*7)]
    })
)
```

Notice that we only specified the `treatment` module, and within that module
only the unnamed code block. Everything else stays as it was in the base
scenario, and all the modules downstream from `treatment` will automatically
pick up the new code when the derived scenarios are run.

### Creating a scenario list

Now, we could run each of the drived scenarios independently, but instead will
will combine them into a list that can be run using a single call:

```{r, eval=F}
sclist <- scenario_list(base_scenario, scenario_fixed_200mg_q2w, scenario_fixed_300mg_q3w)

CTS(sclist)
```

All 3 scenario are execuated.


### Creating a scenario grid

Frequently, we want to explore all possible combinations of different factors.
For this, we can use as scenario grid. Here, we will simulate each of 3 dosing
regimens under 2 different models, one with extimated body weight exponents for
allometric scaling, and the other with fixed (theoretical) exponents.  We can
specify this as follows:

```{r}
scgrid <- scenario_grid(base_scenario,

    parameters = list(
        parameters(name="Estimated allometry"),
        parameters(name="Fixed allometry",
            th = list(CL=0.0194, V1=4.51, Q=0.0435, V2=2.82, CL_WT=0.75, V1_WT=1.00))
    ),

    treatment = list(
        treatment(name="3 mg/kg Q2W", {population[, .(ID, amt=wt*3, tinf=0.5, addl=5, ii=2*24*7)]}),
        treatment(name="200 mg Q2W",  {population[, .(ID, amt=200,  tinf=0.5, addl=5, ii=2*24*7)]}),
        treatment(name="300 mg Q3W",  {population[, .(ID, amt=300,  tinf=0.5, addl=3, ii=3*24*7)]})
    )
)

CTS(scgrid)
```

Each of the 6 scenario will generate the same outputs as the base scenario,
using the specified parameters and treatments. Again, we only need to specify
the elements we wish to override.  Giving the modules discriptive names will be
helpful in the next section when we combine the various scenarios.

### Putting it all together

While outputs from individual scenarios can be important, at the end of the
day, the goal is often to produce a few key tables and/or figures that
summarize and compare a collection of different scenarios. That is the purpose
of the `collate()` function.

```{r}
dat <- collate("endpoints/sim-exposure.csv")
```

What happens is, the outputs generated separately for each scenario are merged
together, along with additional information like the names we gave to each of
the modules.

Hence, we can now produce a nice table to summarize our results across all 6
scenarios:

```{r}
dat <- dat[scenario %in% names(scgrid)]
dat2 <- melt(dat, id.vars=c("parameters", "treatment"), measure.vars=c("cmin", "cmax", "cavg"))

m <- text2mapping("
cmin | C<sub>min,ss</sub> (μg/mL)
cmax | C<sub>max,ss</sub> (μg/mL)
cavg | C<sub>avg,ss</sub> (μg/mL)
")
dat2$variable <- m(dat2$variable)

dat2$parameters <- cf(dat2$parameters)
dat2$treatment  <- cf(dat2$treatment)

label(dat2$parameters) <- "Model"
label(dat2$variable)   <- "Exposure<br/>Metric"

ttt(value ~ parameters + variable | treatment, data=dat2, lab="Regimen<br/>Median [90% PI]",
    render=function(x) with(stats.apply.rounding(stats.default(x)),
        sprintf("%s [%s, %s]", MEDIAN, q05, q95)))
```

## Managing RNG seeds

Because simulations usually involve the generation of random data, storing RNG
seeds is important for reproducibility. By default, each new scenario gets its
own unique seed (which can be overridden) that gets stored with the object.
Each scenario's RNG seed is stored in the `scenarios.csv` table, so it can
easily be reproduced. Still it is a good idea to use `set.seed()` at the start.
Individual modules can also have their own seeds, but by default they do not.
If it were desired to simulate multiple different treatments with the same
population, for instance, then that could be accomplished be having a common
RNG seed for the `population` module.

When a scenario or module has its own RNG seed specified, then the current RNG
state will be saved before running the scenario or module and restored
afterwards.

## TBD

- uncertainty/repetition
- caching
- parallelization

```{r, echo=FALSE}
# Cleanup
unlink(cts_results_path, recursive=T)
```

