---
title: "`ctsframework` : A Modular Framework for Performing Clinical Trial Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`ctsframework` : A Modular Framework for Performing Clinical Trial Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval=TRUE, message=FALSE, warning=FALSE, error=FALSE)
```

```{r, echo=FALSE}
#devtools::load_all()
library(ctsframework)

# Reset all counters
reset_counter(.all=TRUE)

# Create a temporary directory to hold the results
if (file.exists("cts_results")) {
    stop("cts_results already exists and would be overwritten")
}
dir.create("cts_results", FALSE, TRUE)
cts_results_path <- file.path("cts_results", as.Date(Sys.time()))
dir.create(cts_results_path, FALSE, TRUE)
```




# Introduction

One of the core elements of pharmacometrics is to perform clinical trial
simulations (CTS) to help inform decisions in clinical drug development (for
instance, to explore different dosing regimens). The de facto computing
environment for this is R, and there are many excellent packages that can be
leveraged.  However, what tends to happen is that the number of different
"scenarios" that need to be simulated grows exponentially, and as a
consequence the resulting code tends to explode in complexity and become
cumbersome to manage. The purpose of the `ctsframework` package is to make the
CTS process more modular, and to help manage the complexity.

Much as the infamous `ggplot2` package is designed around a "grammar of
graphics", this package offers a "grammar" of semantically named functions that
correspond to different components of a clinical trial simulation.  Designed
for maximum flexibility and convenience. The central object is the scenario.
Specific functions or "modules" correspond to different aspects of CTS which
helps with semantics.


Withing `ctsframework`, a CTS is specified by one or more scenarios. A scenario
is broken down into multiple steps or "modules":

- **parameters**: This module can be used to specify any general parameters
  that are related to a scenario.

- **population**: In most clinical trial simulations, the starting point is
  defining a population of 'virtual patients'. This module is where you would
  specify the distribution of demographic characteristics, implement
  inclusion/exclusion criteria and so on.

- **subject_model**: I don't usually like referring to people as 'subjects',
  but this is the common terminology used in clinical trials. This is the place
  where you could specify subject-level differences, like covariate effects or
  between-subject variability, and individual-level parameters.

- **treatment**: In most clinical trial simulations, the treatment will be of
  primary interest. This module is intended for specifying treatments, which
  can be a dose, formulation etc.

- **pkpd_model**: In pharmacometric applications, there will typically be a
  PK/PD model consisting of by differential equations, which can be specified
  in this module.

- **simulation**: At this point, it's time to carry out the actual simulation.

- **endpoints**: From the simulation output (which could for instance be a
  longitudinal data set), a number of endpoints can be derived, which is what
  this module is intended for.

- **summarization**: This module is intended for summarizing results across
  individuals, such as taking means, medians, specific quantiles, etc. (Note:
  this module can only be used to summarize results within a scenario, not
  across scenarios; the latter is the purpose of the `collate()` function,
  described below).

At each step, the results of the models that appear higher up are available and
can be used in the subsequent computation. By mixing and matching different
modules, it is easy to specify many different scenarios without a lot of
code repetition.

## A complete pharmacometrics example

For this example we will use the following packages, so load them first:

```{r}
library(data.table)
library(mrgsolve)
library(linpk)
library(mappings)
library(ggplot2)
library(table1)
library(ttt)
```

Also, for reproducibility we better set the RNG seed:

```{r}
set.seed(2755139)
```


### Defining a base scenario

```{r}
base_scenario <- scenario(name="Base Scenario",

    parameters = parameters(
        n  = 1000,
        th = list(CL=0.0194, V1=4.51, Q=0.0435, V2=2.82, CL_WT=0.46, V1_WT=0.87),
        om = LTmat(c(0.1, 0.01, 0.1), .names=c("nCL", "nV1"))
    ),

    population = population(
        wtcatfn = cut_mapping(c(0, 60, 100, Inf), c("<60 kg", "60 to <100 kg", ">= 100 kg"), right=TRUE),
        {
            data.table(ID=1:n, wt=exp(rnorm(n, log(74), 0.2)))[, wtcat := wtcatfn(wt)]
        }
    ),

    subject_model = subject_model(
        {
            eta <- generateETA(n, omega=om)

            cbind(population, eta)[, `:=`(
                CL = th$CL * exp(nCL) * (wt/70)^th$CL_WT,
                V1 = th$V1 * exp(nV1) * (wt/70)^th$V1_WT
            )]
        }
    ),

    treatment = treatment(
        {
            population[, .(ID, amt=wt*3, tinf=0.5, addl=5, ii=2*24*7)]
        }
    ),

    pkpd_model = pkpd_model(
        {
            model.code <- '
            [param]   CL=1, V1=1, V2=1, Q=1
            [pkmodel] cmt="CENTRAL PERIPH"
            [table]   capture conc = CENTRAL/V1;
            '

            mcode_cache("2cmt", model.code)
        }
    ),

    simulation = simulation(
        {
            eoi <- realize_addl(as.ev(treatment)) |> as.data.frame() |> with(time + (amt/rate)) |> unique() |> sort()

            sim <- pkpd_model |>
                param(th) |>
                idata_set(subject_model) |>
                ev(as.ev(treatment)) |>
                Req(conc) |>
                update(end=12*24*7, delta=24, add=eoi) |>
                mrgsim()

            simdat <- sim |> as.data.frame() |> as.data.table()

            population[simdat, on="ID"]
        }
    ),

    endpoints = endpoints(
        {
            auc.by.trapezoid <- function(x, y) sum(0.5*diff(x)*(y[-1] + y[-length(y)]))

            simulation[time >= 6*24*7, .(
                cmin=min(conc),
                cmax=max(conc),
                cavg=auc.by.trapezoid(time, conc)/diff(range(time))),
            by=ID][population, on="ID"]
        },

        outputs = csv_output(endpoints, name="sim-exposure")
    ),

    summarization = summarization(
        {
            list(
                overall = simulation[, qi(conc, level=0.9), by=time],
                bywtcat = simulation[, qi(conc, level=0.9), by=.(time, wtcat=wtcatfn(wt))]
            )
        },

        outputs = output_list(
            graphical_output(name="sim-profile-90pctPI-overall", width=7, height=5, {
                ggplot(summarization$overall, aes(x=time, y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (h)", y="Concentration (\U{00B5}g/mL)") +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            }),
            graphical_output(name="sim-profile-90pctPI-bywtcat", width=7, height=5, {
                ggplot(summarization$bywtcat, aes(x=time, y=q0.5, ymin=q0.05, ymax=q0.95)) +
                    labs(x="Time (h)", y="Concentration (\U{00B5}g/mL)") +
                    facet_wrap(~ wtcat) +
                    geom_ribbon(color=NA, fill="blue", alpha=0.1) +
                    geom_line() +
                    theme_bw()
            })
        )
    )
)
```

### Running a scenario

```{r, eval=F}
CTS(base_scenario)
```

After the scenario has finished running, you will find that there is a folder
in the current directory named `cts_results/2023-01-20/Base Scenario`. Within
that folder you will find a subfolder called `summarization` (the same name as
the module) that contains these 2 plots:

```{r, eval=T, echo=F, fig.width=7, fig.height=5}
x <- CTS(base_scenario, .path=cts_results_path)
outputs <- attr(x$scenario$summarization, "outputs", exact=TRUE)
outres <- rlang::eval_tidy(outputs, x)
outres[[1]]
outres[[2]]
```

### Deriving a new scenario from an existing one

```{r, eval=F}
scenario_fixed_200mg_q2w <- derive_scenario(base_scenario,
    treatment = treatment(name="200 mg Q2W", {
        population[, .(ID, amt=200, tinf=0.5, addl=5, ii=2*24*7)]
    })
)

scenario_fixed_300mg_q3w <- derive_scenario(base_scenario,
    treatment = treatment(name="300 mg Q3W", {
        population[, .(ID, amt=300, tinf=0.5, addl=3, ii=3*24*7)]
    })
)
```

### Creating a scenario list

```{r, eval=F}
sclist <- scenario_list(base_scenario, scenario_fixed_200mg_q2w, scenario_fixed_300mg_q3w)

CTS(sclist)
```


### Creating a scenario grid

```{r}
scgrid <- scenario_grid(base_scenario,

    parameters = list(
        parameters(name="Estimated allometry"),
        parameters(name="Fixed allometry",
            th = list(CL=0.0194, V1=4.51, Q=0.0435, V2=2.82, CL_WT=0.75, V1_WT=1.00))
    ),

    treatment = list(
        treatment(name="3 mg/kg Q2W", {population[, .(ID, amt=wt*3, tinf=0.5, addl=5, ii=2*24*7)]}),
        treatment(name="200 mg Q2W",  {population[, .(ID, amt=200,  tinf=0.5, addl=5, ii=2*24*7)]}),
        treatment(name="300 mg Q3W",  {population[, .(ID, amt=300,  tinf=0.5, addl=3, ii=3*24*7)]})
    )
)

CTS(scgrid)
```

Each scenario will generate the same outputs as the base scenario, using the
specified parameters and treatment regimens.

### Putting it all together

While outputs from individual scenarios can be important, at the end of the
day, the goal is often to produce a few key tables and/or figures that
summarize and compare a collection of different scenarios. That is the purpose
of the `collate()` function.

There are 2 ways of doing this:

  1. Using the object returned by the `CTS()` function, which contains multiple
     scenario results in memory.

  2. Reading in the data that was saved to disc during the execution of each
     scenario.


In this example, we will use the second approach.

```{r}
dat <- collate("endpoints/sim-exposure.csv")
```

Now, we can produce a nice table to summarize our results across all 6 scenarios:

```{r}
dat <- dat[scenario %in% names(scgrid)]
dat2 <- melt(dat, id.vars=c("parameters", "treatment"), measure.vars=c("cmin", "cmax", "cavg"))

m <- text2mapping("
cmin | C<sub>min,ss</sub> (μg/mL)
cmax | C<sub>max,ss</sub> (μg/mL)
cavg | C<sub>avg,ss</sub> (μg/mL)
")
dat2$variable <- m(dat2$variable)

dat2$parameters <- cf(dat2$parameters)
dat2$treatment  <- cf(dat2$treatment)

label(dat2$parameters) <- "Model"
label(dat2$variable)   <- "Exposure<br/>Metric"

ttt(value ~ parameters + variable | treatment, data=dat2, lab="Regimen<br/>Median [90% PI]",
    render=function(x) with(stats.apply.rounding(stats.default(x)),
        sprintf("%s [%s, %s]", MEDIAN, q05, q95)))
```

## Managing RNG seeds

Because simulations usually involve the generation of random data, storing RNG
seeds is important for reproducibility. By default, each new scenario gets its
own unique seed (which can be overridden) that gets stored with the object.
Individual modules can also have their own seeds, but by default they do not.
If it were desired to simulate multiple different treatments with the same
population, for instance, then that could be accomplished be having a common
RNG seed for the `population` module.

When a scenario or module has its own RNG seed specified, then the current RNG
state will be saved before running the scenario or module and restored
afterwards.

## TBD

- uncertainty/repetition
- caching
- parallelization

```{r, echo=FALSE}
# Cleanup
unlink(cts_results_path, recursive=T)
unlink("cts_results", recursive=T)
```

